---
title: "Rethinking - Chapter 14"
format: 
  html:
    code-fold: show
toc: true
---
```{r}
#| output: false
#| code-fold: true

library(rethinking)
library(dagitty)
library(tidyverse)
library(cmdstanr)
# rstudioapi::getActiveDocumentContext()$path |> dirname() |> setwd()
```

## 14.0

Consider the simple model:

$$\mu_i = \alpha_{\text{CAFE}_i} + \beta_{\text{CAFE}_i} A_i$$

where $A_i$ is an indicator variable for if we are in the afternoon. If we are modeling wait times are cafes, and we have the above model, it is basically telling us the average wait time for $A_i = 1 $ the afternoon, or $A_i = 0$, the morning. So the intercept ($\alpha$) is the average wait time in the morning, and the slope is the difference to get to the afternoon wait time ($\text{Afternoon} = \alpha + \beta$) 

The thing is, there is probably some correlation between these two wait times. Cafes that are busy in the morning are usually busy in the afternoon. So we can actually model this correlation in our varying intercept/slope model. 

## 14.1 - Varying Slopes by Construction

Let's define some data that we can simulate from:

```{r}
a <- 3.5 # average morning wait time
b <- -1  # average difference between afternoon wait time
sigma_a <- 1 # std dev in intercepts
sigma_b <- 0.5 # std dev in slopes
rho <- -0.7 # correlation between slope and intercept
```

The covariance matrix is defined as:

\begin{bmatrix}
\sigma^2_\alpha & \sigma_\alpha \sigma_\beta \rho \\
\sigma_\alpha \sigma_\beta \rho & \sigma^2_\beta \\
\end{bmatrix}

```{r}
cov_ab <- sigma_a * sigma_b * rho
Sigma <- matrix(c(sigma_a^2, cov_ab, cov_ab, sigma_b^2), ncol=2)

N_cafes <- 20
set.seed(5)

vary_effects <- MASS::mvrnorm(N_cafes, c(a,b), Sigma)

a_cafe <- vary_effects[,1]
b_cafe <- vary_effects[,2]

plot(a_cafe, b_cafe, col=rangi2, xlab='Intercepts', ylab='slopes')

# overlay population distribution
for (l in c(0.1, 0.3, 0.5, 0.8, 0.99)){
  lines(ellipse::ellipse(Sigma, centre=c(a,b), level=l),
        col=col.alpha('black', 0.2))
}
```

Quick detour into the def of covariance:

We all know $\text{cov}(x,y) = E(xy) - E(x)E(y)$. So this is really just the difference between the average product and the product of the averages. We could of course center $x$ and $y$ such that $\text{cov}(x,y) = E(xy)$. We can do the same with the variances too. Now, where does $\rho$ come from? $\rho$ is the ratio of the covariance to the maximum that the covariance could theoretically ever be: $\sqrt{\text{var}(x) \text{var}(y)}$.

Where does this come from? Well if $y = \beta x$, for some scalar $\beta$ (can be anything), then the covariance is at its maximum. Think about it this way, $x$ would be a **perfect** predictor of $y$. There is no noise, they aren't independent, in fact, they are as dependent as possible!

So if $y=  \beta x$ and we center, then $\text{cov}(x,y) = E(xy) = \beta E(x^2)$ and $\text{var}(x) \text{var}(y) = E(x^2) E(y^2) = \beta^2E(x^2)^2$. Then $\text{cov}(x,y) = \sqrt{\text{var}(x) \text{var}(y)}$ - the max covariance. So we define $\rho = \text{cov}(x,y) / \sqrt{\text{var}(x) \text{var}(y)}$

```{r}
a1 <- rnorm(100)
a2 <- a1 * 1.2
all.equal(cov(a1,a2), sqrt(var(a1)*var(a2)))
```

Now let's have the robot go and visit cafes 10 times each - 5 in the morning and 5 in the afternoon:

```{r}
set.seed(22)
N_visits <- 10
afternoon <- rep(0:1, N_visits*N_cafes/2)
cafe_id <- rep(1:N_cafes, each=N_visits)

mu <- a_cafe[cafe_id] + b_cafe[cafe_id]*afternoon
sigma <- 0.5
wait <- rnorm(length(mu), mu, sigma)
d <- data.frame(cafe=cafe_id, afternoon=afternoon, wait=wait)
str(d)
```

Let's now build out the full model to estimate:

\begin{align}
W_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{\text{CAFE}_i} + \beta_{\text{CAFE}_i} A_i \\ \\
\begin{bmatrix}
\alpha_{\text{CAFE}} \\
\beta_{\text{CAFE}}
\end{bmatrix}
 &\sim \text{MVNormal}(\begin{bmatrix}
\alpha \\
\beta
\end{bmatrix}, S ) \\ \\
S &= 
\begin{pmatrix}
\sigma_\alpha & 0 \\
0 & \sigma_\beta
\end{pmatrix}
R 
\begin{pmatrix}
\sigma_\alpha & 0 \\
0 & \sigma_\beta
\end{pmatrix}
\end{align}

For the correlation matrix $R$, we'll use the typical LKJcorr prior. Specifically one where $\eta = 2$. Recall that is $\eta = 1$, we are completely flat and allow equal weighting to all correlations between -1 and 1. For $\eta = 2$, we are skeptical about large correlations:

```{r}
for (e in c(1,2,4)){
  R <- rlkjcorr(1e4, K=2, eta=e)
  if (e == 1) dens(R[,1,2], xlab='correlation', ylim=c(0,1.3)) else dens(R[,1,2], add=T)
}
```

Let's now fit the model:

```{r}
# m14.1 <- ulam(alist(
#   wait ~ normal(mu, sigma),
#   mu <- a_cafe[cafe] + b_cafe[cafe] * afternoon,
#   c(a_cafe, b_cafe) ~ multi_normal(c(a,b), Rho, sigma_cafe),
#   a ~ normal(5,2),
#   b ~ normal(-1,0.5),
#   sigma_cafe ~ exponential(1),
#   sigma ~ exponential(1),
#   Rho ~ lkj_corr(2)
# ), data=d, chains=4, cores=4)
```

```{r}
#| output: false

mod <- cmdstan_model('mod_14.1.stan')

fit <- mod$sample(
  data = list(
    N       = nrow(d),
    N_cafe  = max(d$cafe),
    cafe    = d$cafe,
    afternoon = d$afternoon,
    wait    = d$wait
  ),
  chains = 4,
  parallel_chains = 4,
  refresh = 500,
  iter_warmup = 1e3, iter_sampling=1e3
)
```

```{r}
fit$summary('Rho') |> print(n=100)
```

Let's see what the prior tension is for the correlation matrix:


```{r}
mod_rho <- fit$draws('Rho[2,1]', format='df') %>% pull('Rho[2,1]')
dens(mod_rho)
R <- rlkjcorr(1e4, K=2, eta=2)
dens(R[,1,2], add=T, lty=2)
```

Now let's plot the estimates and their shrinkage:

```{r}
full_draws <- fit$draws(format='df') %>% select(-starts_with('.'))
```
```{r}
# compute unpooled estimates directly from data
a1 <- sapply( 1:N_cafes , function(i) mean(wait[cafe_id==i & afternoon==0]) )
b1 <- sapply( 1:N_cafes , function(i) mean(wait[cafe_id==i & afternoon==1]) ) - a1

# extract posterior means of partially pooled estimates


a2 <- full_draws %>% select(starts_with('ab_cafe')) %>% select(ends_with('1]')) %>% colMeans()
b2 <- full_draws %>% select(starts_with('ab_cafe'))  %>% select(ends_with('2]')) %>% colMeans()
# plot both and connect with lines
plot( a1 , b1 , xlab="intercept" , ylab="slope" ,
  pch=16 , col=rangi2 , ylim=c( min(b1)-0.1 , max(b1)+0.1 ) ,
  xlim=c( min(a1)-0.1 , max(a1)+0.1 ) )
points( a2 , b2 , pch=1 )

for ( i in 1:N_cafes ) lines( c(a1[i],a2[i]) , c(b1[i],b2[i]))

mu_est <- full_draws %>% select(starts_with('mu_ab')) %>% colMeans()
rho_est <- full_draws %>% select(starts_with('Rho[1,2]')) %>% unlist() %>% mean()
sig_cafe <- full_draws %>% select(starts_with('sigma_cafe')) %>% colMeans()
Sigma <- diag(sig_cafe) %*% matrix(c(1, rho_est, rho_est, 1), ncol=2) %*% diag(sig_cafe)

for (l in c(0.1, 0.3, 0.5, 0.8, 0.99)){
  lines(ellipse::ellipse(Sigma, centre=mu_est, level=l),
        col=col.alpha('black', 0.2))
}
```

The blue points in the plot are the observed values and the open points are the shruken ones. Now lets display it on the outcome scale of wait times:


```{r}
# convert varying effects to waiting times
wait_morning_1 <- (a1)
wait_afternoon_1 <- (a1 + b1)
wait_morning_2 <- (a2)
wait_afternoon_2 <- (a2 + b2)

# plot both and connect with lines
plot( wait_morning_1 , wait_afternoon_1 , xlab="morning wait" ,
  ylab="afternoon wait" , pch=16 , col=rangi2 ,
  ylim=c( min(wait_afternoon_1)-0.1 , max(wait_afternoon_1)+0.1 ) ,
  xlim=c( min(wait_morning_1)-0.1 , max(wait_morning_1)+0.1 ) )
points( wait_morning_2 , wait_afternoon_2 , pch=1 )

for ( i in 1:N_cafes )
  lines( c(wait_morning_1[i],wait_morning_2[i]) ,
         c(wait_afternoon_1[i],wait_afternoon_2[i]) )
abline( a=0 , b=1 , lty=2 )

# we'll simulate the wait times to get contours to keep the data generating persepctive
  # To truely undertand the uncertainty we should probably sample from the hyperparams too
  # But we'll keep those fixed to keep it clean

v <- MASS::mvrnorm(1e4, mu_est, Sigma)
v[,2] <- v[,1] + v[,2] #calculate afternoon wait time
Sigma2 <- cov(v)
mu_est2 <- colMeans(v)

for (l in c(0.1, 0.3, 0.5, 0.8, 0.99)){
  lines(ellipse::ellipse(Sigma2, centre=mu_est2, level=l),
        col=col.alpha('black', 0.2))
}
```

The thing to catch here is that the shrinkage on the parameter scale has naturally made its way into shrinkage on the outcome scale.

## 14.2 - Advanced Varying Slopes

Recall the chimp problem from lat chapter. We had different actors in different blocks with different treatments. Lots of effects to keep track of:

\begin{align}
L_i &\sim \text{Binomial}(1, p_i) \\
\text{logit}(p_i) &= \gamma_{\text{TID}_i} + \alpha_{\text{ACTOR}_i,\text{TID}_i} + \beta_{\text{BLOCK}_i,\text{TID}_i}
\end{align}

This is basicallly a big interaction effects model. We are allowing the effect of each actor to varying by treatment, within each block. So there is a total of $4 + 7\cdot 4 + 6 \cdot 4 = 56$ params. 


I think it makes sense to start the actor to aid in the understanding. For each actor, we want to estimate some mean vector that perturbs the intercept specifcally for them. This mean vector is drawn from a super population of actors that have similar mean vectors. We can use this shard information since some actors have more observations in some treatments then others. It feels weird because we are leaving behind the nice intercept/slope defintion, but it is all the same really. We have some effect, per treatment, that varies on the actor level. Perfect for a multi-level model for regularization.

For the actor, the hyper-prior looks something like:
$$
\begin{bmatrix}
\alpha_{j,1} \\ \alpha_{j,2} \\ \alpha_{j,3} \\ \alpha_{j,4}
\end{bmatrix} \sim \text{MVNormal} \left (
\begin{bmatrix}
0 \\ 0 \\ 0 \\ 0
\end{bmatrix},
S_{\text{ACTOR}}
  \right)
$$

```{r}
data(chimpanzees)
d <- chimpanzees
d$block_id <- d$block
d$treatment <- 1L + d$prosoc_left + 2L*d$condition
dat <- list(
  L = d$pulled_left,
  tid = d$treatment,
  actor = d$actor,
  block_id = as.integer(d$block_id) )

dat %>% as.data.frame() %>% apply(2, n_distinct)

# m14.2 <- ulam(alist(
#   L ~ dbinom(1,p),
#   logit(p) <- g[tid] + alpha[actor,tid] + beta[block_id,tid],
#   # adaptive priors
#   vector[4]:alpha[actor] ~ multi_normal(0,Rho_actor,sigma_actor),
#   vector[4]:beta[block_id] ~ multi_normal(0,Rho_block,sigma_block),
#   # fixed priors
#   g[tid] ~ dnorm(0,1),
#   sigma_actor ~ dexp(1),
#   Rho_actor ~ dlkjcorr(4),
#   sigma_block ~ dexp(1),
#   Rho_block ~ dlkjcorr(4)
#   ) , data=dat , chains=4 , cores=4 )
```

We'll of course use Stan to fit:


```{r}
#| output: false
dat$N <- d |> nrow()
dat$N_actor <- dat$actor |> n_distinct()
dat$N_block <- dat$block_id |> n_distinct()
dat$N_tid <- dat$tid |> n_distinct()


mod <- cmdstan_model('mod_14.2.stan')
fit <- mod$sample(dat, chains=4, iter_warmup=1e3, iter_sampling=1e3, refresh=500, parallel_chains=4)
```

Richie talks about reparm which we have already done in the Stan code. The reparam trick comes from the fact that we are declaring a standard normal matrix and then mutliplying by the covariance matrix to get back to the "original" distribution. Again, effectively decoupling the mean from the variance.

Now, recall that we had roughly 56 params, plus the 8 standard deviations and 12 correlation params equals 76. If we look at the p_loo, we'll see it is much lower:


```{r}
f_loo <- fit$draws('log_lik', format='matrix') |> loo::loo()
f_loo
```

This is because if the high degree of shrinkage the model imposes. Seen a different way, look at the high sigma values:

```{r}
fit$summary(c('sigma_actor', 'sigma_block'))
```

## 14.3 - Instruments and Causal Designs

Introduce IVs, they are a new tool in the toolkit for identifying causal effects. Let's we want education's effect on wages: $E \rightarrow W$. But there is a backdoor path with a confounder $U$: $E \leftarrow U \rightarrow W$. Now, we if have a variable $Q$ that acts as a natural experiment on $E$, then we can use it to determine the causal effect on $W$. But it must statisfy the follow:

- Independent of $U$, $(Q \perp\!\!\!\perp U)$
- Not independent of $E$
- $Q$ cannot influence $W$, except though $E$ (exclusion criteria)

Naively, we might think we can just include $Q$ in the model, but look at the path: $Q \rightarrow E \leftarrow U \rightarrow W$. Since are conditioning on $E$, it becomes a collider and $E$ can be even more confounded. In this case, we would call $Q$ a bias amplifier. 

Let's simulate some data before we get to the nitty-gritty. We'll have $Q$ be birth quarter which used to correspond to the amount of schooling a student might get:

```{r}
N <- 500
U_sim <- rnorm(N)
Q_sim <- sample(1:4, size=N, replace=T)
E_sim <- rnorm(N, U_sim + Q_sim)
W_sim <- rnorm(N, U_sim + 0*E_sim)
dat_sim <- list(
  W = standardize(W_sim),
  E = standardize(E_sim),
  Q = standardize(Q_sim)
)
```
For the sake of showing effects, we have set the effect of education on wages to be zero. But the instrument does influence education and education alone. It is also true that a higher value for $Q$ yields a higher value for $E$. Let's try three different models. First, a naive regression:

```{r}
m14.4 <- quap(alist(
  W ~ dnorm(mu, sigma),
  mu <- aW + bEW*E,
  aW ~ dnorm(0, 0.2),
  bEW ~ dnorm(0, 0.5),
  sigma ~ dexp(1)
), data=dat_sim)

precis(m14.4)
```

Ordinary confounding. Let's now add $Q$ to the model:

```{r}
m14.5 <- quap(alist(
  W ~ dnorm(mu, sigma),
  mu <- aW + bEW*E + bQW*Q,
  aW ~ dnorm(0, 0.2),
  bEW ~ dnorm(0, 0.5),
  bQW ~ dnorm(0, 0.5),
  sigma ~ dexp(1)
), data=dat_sim)

precis(m14.5)
```

Wow - even worse. 

To get this correctly, we need to model the data generating process explicitly. That is, a separate model for education, and one for wages.

Wages looks like:

\begin{align}
W_i &\sim \text{Normal}(\mu_{W,i}, \sigma_W) \\
\mu_{W,i} &= \alpha_W + \beta_{EW} E_i + U_i \\
\end{align}

and education looks like:

\begin{align}
E_i &\sim \text{Normal}(\mu_{E,i}, \sigma_E) \\
\mu_{E,i} &= \alpha_E + \beta_{QE} Q_i + U_i \\
\end{align}

Finally, we have $Q_i \sim \text{Categorical}(0.25, 0.25,0.25,0.25)$ and $U_i\sim \text{Normal}(0,1)$ for the instrument and confounder, respectively. 

To handle the confounder, we will average over it and model the covariance between $W$ and $E$:

\begin{align}
\begin{pmatrix}
W_i \\ E_i
\end{pmatrix}
& \sim \text{MVNormal}([\mu_{W,i}, \mu_{E,i}], S) \\
\mu_{W,i} &= \alpha_W + \beta_{EW} E_i \\
\mu_{E,i} &= \alpha_E + \beta_{QE} Q_i
\end{align}

It might feel weird to see $E$ as both an observed variable and covariate for the $W$ regression equation, but there is nothing illegal aboout this. Also, the covariance matrix here is to capture the correlation that might be present in the residuals of predicting $E$ and $W$, through $U$. You can think of this as: if the error on $E$ increases, so will the error on $W$. Why? Because that most likely correlates to a high $U$ that contributed to that individual residual.

Let's now model it:

```{r}
# m14.6 <- ulam(
#   alist(
#     c(W,E) ~ multi_normal( c(muW,muE) , Rho , Sigma ),
#     muW <- aW + bEW*E,
#     muE <- aE + bQE*Q,
#     c(aW,aE) ~ normal( 0 , 0.2 ),
#     c(bEW,bQE) ~ normal( 0 , 0.5 ),
#     Rho ~ lkj_corr( 2 ),
#     Sigma ~ exponential( 1 )
#   ), data=dat_sim , chains=4 , cores=4 )
# 
# precis( m14.6 , depth=3 )
```

```{r}
#| output: false

mod <- cmdstan_model('mod_14.6.stan')
dat_sim$N <- dat_sim$Q |> length()
fit <- mod$sample(dat_sim, chains=4, iter_warmup=1e3, iter_sampling=1e3, refresh=500, parallel_chains=4)
```
```{r}
fit$summary(c('aE', 'aW', 'bQE', 'bEW', 'Rho[1,2]'))
```

Notice that $\beta_{EW}$ is now straddling zero. 

Let's try with some different simulation params:

```{r}
#| output: false

N <- 500
U_sim <- rnorm(N)
Q_sim <- sample(1:4, size=N, replace=T)
E_sim <- rnorm(N, U_sim + Q_sim)
W_sim <- rnorm(N, -U_sim + 0.2*E_sim)
dat_sim <- list(
  W = standardize(W_sim),
  E = standardize(E_sim),
  Q = standardize(Q_sim),
  N = N
)

fit <- mod$sample(dat_sim, chains=4, iter_warmup=1e3, iter_sampling=1e3, refresh=500, parallel_chains=4)
```

```{r}
fit$summary(c('aE', 'aW', 'bQE', 'bEW', 'Rho[1,2]'))
```

Notice now that there is a negative correlation between the residuals. This is because $U$ positively influences $E$, but negatively influences $W$. 

We can also use `dagitty` to help us find instrument variables, if they are present in the DAG:


```{r}
dagIV <- dagitty("dag{Q -> E <- U -> W <- E}")
instrumentalVariables(dagIV, exposure="E", outcome='W')
```

Some final thoughts:

- People sometimes use 2SLS (2-stage least squares) for estimating IV
  - Process where we regress $Q$ onto $X$, then take the predicted values from that equation and feed them into the regression of $\hat X$ onto $Y$
  - This makes some strong assumptions and requires large sample sizes to get reliable estaimtes (check F-stat from first stage)
- Use mediator variables to get causal estimates
  - consider the DAG $X \rightarrow Z \rightarrow Y \leftarrow U \rightarrow X$
  - Clearly $Z$ is un-confounded by $U$, therefore we can use it to get the causal estimate of $X$ on $Y$
- Regresssion Discontinuity Design
  - Consider a decision that is based on a continuous GPA
  - There will be students who just fall above and those that just fall below this "arbitrary" cut point
  - We can use this to estimate the causal effect by fitting two separate regressions - one before and one after the decision point

## 14.4 - Social relations and correlated varying effects

Consider the `kl_dyads` and `kl_households` datasets. The first is a dataset of households from Nicaragua where these households have given gifts between one another. The variables `giftsAB` and `giftsBA` are the gifts in each direction within each dyad. A dyad here is a pair of households. 

```{r}
data(KosterLeckie)
```

The model for $A \rightarrow B$ will look like such:

\begin{align}
y_{A \rightarrow B} & \sim \text{Poisson}(\lambda_{AB}) \\ 
\log \lambda_{AB} &= \alpha + g_A + r_B + d_{AB}
\end{align}

Where $g_A$ is the household $A$ specific tendancy to give, $r_B$ is the household specific tendancy to receive, and $d_{AB}$ is the dyad specific effect of $A$ to $B$.

There is of course another model for $B\rightarrow A$

\begin{align}
y_{B \rightarrow A} & \sim \text{Poisson}(\lambda_{BA}) \\ 
\log \lambda_{BA} &= \alpha + g_B + r_A + d_{BA}
\end{align}

We will use a bivariate normal to represent the adaptive prior on the $g$ and $r$ effects:

$$
\begin{pmatrix} g_i\\ r_i \end{pmatrix} 
\sim 
\text{MVNormal}(0, \begin{pmatrix} \sigma_g^2 & \sigma_g \sigma_r \rho_{gr} \\ \sigma_g \sigma_r \rho_{gr} & \sigma_r^2  \end{pmatrix})
$$

We'll do something similar for the dyad effects:

$$
\begin{pmatrix} d_{ij}\\ d_{ji} \end{pmatrix} 
\sim 
\text{MVNormal}(0, \begin{pmatrix} \sigma_d^2 & \sigma_d^2 \rho_{d} \\ \sigma_d^2 \rho_{d}  & \sigma_d^2  \end{pmatrix})
$$

Notice that we are using the same variance $\sigma_d$ here. That is because the labels in each dyad are arbitrary, and it should matter if it is 1 to 2 or 2 to 1. That is, it isn't meaningful which household comes first. But we do want to estimate the correlation.To interpret: if $\rho_d$ is positive, then if one household gives more within a dyad, then so too does the other.

Let's go ahead and run our Stan model:


```{r}
#| output: false
kl_data <- list(
  n = nrow(kl_dyads),
  n_households = max(kl_dyads$hidB),
  did = kl_dyads$did,
  hidA = kl_dyads$hidA,
  hidB = kl_dyads$hidB,
  giftsAB = kl_dyads$giftsAB,
  giftsBA = kl_dyads$giftsBA
)

mod <- cmdstan_model('mod_14.7.stan')
fit <- mod$sample(kl_data, chains=4, iter_warmup=1e3, iter_sampling=1e3, refresh=500, parallel_chains=4)
```


```{r}
fit$summary(c('Rho_gr[1,2]', 'Rho_d[1,2]', 'sigma_gr'))
```

The negative correlation of `Rho_gr` implies the individuals that give more, tend to receive less. It also looks like, based on `sigma_gr`, that rates of giving are more variable than rates of receiving.

Let's plot the receiving effects so we can see the covariance structure:

```{r}
sub_draws <- fit$draws(format='df') %>% select(-starts_with('.')) %>% select('a', starts_with('gr'))
n_hh <- 25
g <- matrix(nrow=nrow(sub_draws), ncol=n_hh)
r <- matrix(nrow=nrow(sub_draws), ncol=n_hh)
for (i in 1:n_hh){
  g[,i] <- sub_draws$a + sub_draws %>% pull(paste0('gr[', i, ',1]'))
  r[,i] <- sub_draws$a + sub_draws %>% pull(paste0('gr[', i, ',2]'))
}
Eg_mu <- apply( exp(g) , 2 , mean )
Er_mu <- apply( exp(r) , 2 , mean )
```
We can observe the uncertainty for the first household:

```{r}
plot(exp(g[,1]), exp(r[,1]), pch='.', xlab='Giving', ylab='Receiving', main='HH 1')
```

Let's go ahead and visualize the uncertainty contours for each of our households we've pulled:

```{r}
plot( NULL , xlim=c(0,8.6) , ylim=c(0,8.6) , xlab="generalized giving" ,
  ylab="generalized receiving" , lwd=1.5 )
abline(a=0,b=1,lty=2)

library(ellipse)
for ( i in 1:25 ) {
  Sigma <- cov( cbind( g[,i] , r[,i] ) )
  Mu <- c( mean(g[,i]) , mean(r[,i]) )
  for ( l in c(0.5) ) {
    el <- ellipse( Sigma , centre=Mu , level=l )
    lines( exp(el) , col=col.alpha("black",0.5) )
  }
}
points(Eg_mu ,Er_mu , pch=21, bg="white",lwd=1.5)
```

Note that every ellipse here is the uncertainty around the individual HH's giving and receiving parameter. The dashed line shows where giving and receiving would be equal. Again, this is not the population distribution, this a HHs specific effect.

Let's look at the dyad specific effects. From the `Rho_d[1,2]`, we know that there is roughly a 0.88 correlation for the population distribution

```{r}
dy1 <- fit$draws(format='df') %>% select(starts_with('d')) %>% select(ends_with(',1]')) %>% colMeans()
dy2 <- fit$draws(format='df') %>% select(starts_with('d')) %>% select(ends_with(',2]')) %>% colMeans()
plot(dy1, dy2)
abline(h=0,lty=2)
abline(v=0, lty=2)
abline(a=0,b=1,lty=2)
```

Just a quick note on the dyads here. The model implicitly learns that d[1,1] encodes $A \rightarrow B$ since it always shows up with $g_A$ and $r_B$ in the model equation. It feels weird that we are making the diagonal of the covariance matrix to be equal, but this is really just a structural constraint that we have no reason to believe that $A \rightarrow B$ should be any stronger than $B \rightarrow A$ in general. The model still learns the direction and it still creates separate parameters for $d_{AB}$ and $d_{BA}$. It just feels a little different since we constrain the population distribution to hold their marginal variances to the same value. But, the actual parameters are free to vary. Plus, it may feel weird since we are using the same unique dyad coding for $A \rightarrow B$ and $B \rightarrow A$ (both directions), making it feel like there should be some sort of data relation between the two. But its really the model that separates the two implicitly.

## 14.5 - Continuous categories and the Gaussian process

It's a different mindset, but we can think of a Gaussian processes as continuous groupings instead of discrete objects, where there is some sense of distance (not always physical) between objects.

Let's go back to the Oceanic Islander example data from before. Clearly, there is some notion of distance here, and it has influence on the different tools that a society might have.

```{r}
data("islandsDistMatrix")
Dmat <- islandsDistMatrix
colnames(Dmat) <- c("Ml","Ti","SC","Ya","Fi","Tr","Ch","Mn","To","Ha")
round(Dmat,1)
```

We'll use the distance matrix above to measure a similarity in technology exposure.

Recall the original model from ch 11. Note that the lambda structure is domain specific. $\alpha$ here acts as an intercept:

\begin{align}
T_i &\sim \text{Poisson}(\lambda_i) \\
\lambda_i &= \alpha P_i^\beta / \gamma
\end{align}

We now want to add a varying intercept parameter that varies by `society`:

\begin{align}
T_i &\sim \text{Poisson}(\lambda_i) \\
\lambda_i &= \exp(k_{\text{society}[i]}) P_i^\beta / \gamma
\end{align}

where:

\begin{align}
\begin{bmatrix} k_1 \\ k_2 \\ k_3 \\ \vdots \\ k_{10} \end{bmatrix} &\sim \text{MVNormal}(\vec{0},K) \\
K_{ij} &= \eta^2 \exp(-\rho^2 D^2_{ij}) + \delta_{ij}\sigma^2
\end{align}

This is the heart of the GP - the multivariate prior for the intercepts. Let's break down the equation for $K$:

- $\eta^2$ is the maximum that the covariance is allowed to be between any two societies $i$ and $j$
- $\delta_{ij}$ is just the classic identity function that returns 1 if $i=j$ and 0 otherwise
- $\sigma^2$ is the extra variance that might need to be present because the distance function is 0 for that case
  - Basically just quantifying the noise for the $i=j$ case
  - We don't need it for this model though since we only have one observation per society
- $\exp(-\rho^2 D^2_{ij})$ is the actual distance function
  - Basically the Gaussian kernel 
  - $\rho$ here is analagous to the Gaussian precision. If it is large, then the covariance declines rapidly
  - We square because of convention, but also because it allows for covariance to drop quicker as distance increases


We'll plot the square and non-square below. Notice that it falls slower, but as the distance increase it begins to drop very quickly.

```{r}
curve(exp(-x), 0, 4, lty=2)
curve(exp(-x^2), add=T, 0, 4)
```

The last thing to do is define priors on $\rho^2$ and $\eta^2$, where both must be positive:

\begin{align}
\eta^2 &\sim \text{Exponential}(2) \\ 
\rho^2 &\sim \text{Exponential}(0.5)
\end{align}

Let's now go ahead and fit the model:

```{r}
#| output: false

data(Kline2)# loadtheordinarydata,nowwithcoordinates
d<-Kline2
d$society<-1:10 #index observations

dat_list<- list(
  T=d$total_tools,
  P=d$population,
  society=d$society,
  Dmat=islandsDistMatrix,
  n_societies = length(d$society)
)

mod <- cmdstan_model('mod_14.8.stan')
fit <- mod$sample(dat_list, chains=4, iter_warmup=1e3, iter_sampling=1e3, refresh=500, parallel_chains=4)
```
```{r}
fit$summary(c('k'))
```

Here's another way to look at GP - they are a prior over the function space. What function? Well, the function that maps distance to some sort of linear effect in our link function. So for different parameters that govern the GP, we have essentially a population distribution of functions that is over the possible distance functions.

Let's first see what the covariance function looks like. This is not a function space, but instead the covariance space that implies a function space. The function space is just drawing function functions from the same `x_seq`, but instead plugging in the covariance space with a mean 0:

```{r}
etasq <- fit$draws('etasq', format='df')$etasq
rhosq <- fit$draws('rhosq', format='df')$rhosq

plot( NULL , xlab="distance (thousand km)" , ylab="covariance" , xlim=c(0,10) , ylim=c(0,0.6), main='Gaussian Process Posterior' )
x_seq <- seq(0, 10, length.out=100)
pmcov <- vapply(x_seq, \(x) etasq * exp(-rhosq*x^2), numeric(length(etasq)))
pmcov_mu <- colMeans(pmcov)

lines( x_seq , pmcov_mu , lwd=2 )

# plot 50 functions sampled from posterior
for (i in 1:50){
  curve( etasq[i]*exp(-rhosq[i]*x^2) , add=TRUE ,
      col=col.alpha("black",0.3) )
}
```

So it looks like the covariance function decays rapidly once we get out past 4,000 km. Let's show the implied correlation between societies from this GP:

```{r}
#compute posteriormediancovarianceamongsocieties
K <- matrix(0,nrow=10,ncol=10)
for( i in 1:10)
  for(j in 1:10 )
    K[i,j]<-median(etasq) * exp(-median(rhosq)*islandsDistMatrix[i,j]^2)

diag(K)<-median(etasq) + 0.01

# convert to correlation matrix
Rho<-round( cov2cor(K),2)
# add row/col names for convenience
colnames(Rho)<-c("Ml","Ti","SC","Ya","Fi","Tr","Ch","Mn","To","Ha")
rownames(Rho)<-colnames(Rho)
Rho
```

Note that the societies in the upper left are all highly correlated as they are very close to each other. Hawaii on the other hand is quite far from other islands.

Let's plot these islands:

```{r}
# scale point size to logpop
psize <- d$logpop / max(d$logpop)
psize <- exp(psize*1.5)-2
# plot raw data and labels
plot( d$lon2 , d$lat , xlab="longitude" , ylab="latitude" ,
  col=rangi2 , cex=psize , pch=16 , xlim=c(-50,30) )
labels <- as.character(d$culture)
text( d$lon2 , d$lat , labels=labels , cex=0.7 , pos=c(2,4,3,3,4,1,3,2,4,2) )

# overlay lines shaded by Rho
for( i in 1:10 )
  for ( j in 1:10 )
    if ( i < j )
      lines( c( d$lon2[i],d$lon2[j] ) , c( d$lat[i],d$lat[j] ) ,
            lwd=2 , col=col.alpha("black",Rho[i,j]^2) )
```

The line colors are the strength of the correlations. Darker colors indicate stronger correlations.
