---
title: "Chapter 3"
output: html_document
---

# Ch 3

### 3.1

```{r}
library('lobstr')

x <- c(1,2,3)
```
It is not read as "object named x containing values 1,2, and 3". Instead, think of it as creating an object containing 1,2, and 3 and binding that object to a name, x. 

If I instead write 

```{r, echo=T}
y <- x
obj_addr(x)
obj_addr(y)
```
Then both x and y are bound to the same object.

Consider the following:
```{r, echo=T}
x <- c(1,2,3)
y <- x
c(obj_addr(x),obj_addr(y))
y[[3]] <- 4
c(obj_addr(x),obj_addr(y))
x
```
Noticed that x stayed that same. This is called "copy-on-modify". Initially, y was defined to the same object address as x, but changing the third value in the object of y caused a copy and then modify of x.

Consider the following for lists:
```{r}
l1 <- list(1,2,3)
l2 <- l1
l2[[3]] <- 4
ref(l1,l2)
```
Consider the following for dataframes. If a column is modified, only the column needs to be copy and modified. If a row is modified, a whole new dataframe is created.

```{r}
d1 <- data.frame(x=c(1,2,3), y=c(4,5,6))
d2 <- d1
d2[,2] <- d2[,2]*2
ref(d1,d2)
```
Character vectors are weird. You can think of a character in a vector as belonging to a **global** character pool.

```{r}
x <- c('a', 'a', 'abc', 'f')
ref(x, character = T)
```

Modify and copy with dataframes:
```{r}
x <- data.frame(matrix(runif(5*1e4), ncol=5))
medians <- vapply(x, median, numeric(1))
cat(tracemem(x),'\n')
for (i in seq_along(medians)){
  x[[i]] <- x[[i]] - medians[[i]]
}
untracemem(x)
```
Notice that each loop makes 3 copies of x! This is because "[[.data.frame" is a regular function the increments the reference count of x. We can reduce the number of copies by using a list instead:

```{r}
x <- data.frame(matrix(runif(5*1e4), ncol=5))
medians <- vapply(x, median, numeric(1))
y <- as.list(x)
cat(tracemem(y),'\n')
for (i in seq_along(medians)){
  y[[i]] <- y[[i]] - medians[[i]]
}
untracemem(y)
```
